Resolvable Ambiguity

### 2020-08-12 paper/paper.tex
Pre repository split snapshot

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{cite}
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
%% \usepackage{subcaption} %% For complex figures with subfigures/subcaptions
%%                         %% http://ctan.org/pkg/subcaption
\usepackage{subfig}
\usepackage{wrapfig}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,automata,calc}
\usepackage{tikz-qtree} % Used for syntax trees
\usepackage{tikz-cd} % Used for commutative diagrams
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

@@

\title{Resolvable Ambiguity\thanks{This project is financially supported by the Swedish Foundation for Strategic Research (FFL15-0032)}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Viktor Palmkvist\inst{1} \and
Elias Castegren\inst{1} \and
Philipp Haller\inst{1} \and
David Broman\inst{1}}
%% \author{Viktor Palmkvist\inst{1}\orcidID{0000-0003-0669-4085} \and
%% Elias Castegren\inst{1}\orcidID{0000-0003-4918-6582} \and
%% Philipp Haller\inst{1}\orcidID{0000-0002-2659-5271} \and
%% David Broman\inst{1}\orcidID{0000-0001-8457-4105}}
%
\authorrunning{V. Palmkvist et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{KTH Royal Institute of Technology, 100 44 Stockholm, Sweden
\email{\{vipa,eliasca,dbro,phaller\}@kth.se}%\and
%% Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
%% \email{lncs@springer.com}\\
%% \url{http://www.springer.com/gp/computer-science/lncs} \and
%% ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
%% \email{\{abc,lncs\}@uni-heidelberg.de}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
A common standpoint when designing the syntax of programming languages is that
the grammar definition has to be unambiguous.  However, requiring up front unambiguous grammars
can force language designers to make more or less arbitrary choices to disambiguate
the language. In this paper, we depart from the traditional view of
unambiguous grammar design, and enable the detection of ambiguities to be delayed until
parse time, allowing the user of the language to perform the disambiguation. A
natural decision problem follows: given a language definition, can a
user always disambiguate an ambiguous program?  We introduce and
formalize this fundamental problem---called the \emph{resolvable
  ambiguity problem}---and divide it into separate static and dynamic
resolvability problems. We provide solutions to the static problem
for a restricted language class and sketch proofs of soundness and completeness. We also provide a sound
and complete solution to the dynamic problem for a much less
restricted class of languages.
%
The approach is evaluated through two separate case studies,
covering both a large existing programming language, and the
composability of domain-specific languages.
\keywords{Syntax \and Ambiguity \and Grammars \and Parsing \and Domain-specific languages}
\end{abstract}
%
%
%

@@

% Motivate why unambiguous parsing is a good (and important) idea.
Ever since the early 60s, it has been known that determining whether a context-free
grammar is ambiguous is undecidable~\cite{cantorAmbiguityProblemBackus1962}. As a consequence,
a large number of restricted grammars have been developed to guarantee
that language definitions are unambiguous. This traditional view of
only allowing unambiguous grammars has been taken for granted as the
only truth: \emph{the} way of how the syntax of a language must be
defined ~\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}.
%
However, our recent work~\cite{palmkvistCreatingDomainSpecificLanguages2019} suggests that
carefully designed ambiguous syntax definitions can be
preferable compared to completely unambiguous
definitions.

@@

In this paper we describe a syntax definition formalism where ambiguities can be resolved by grouping parentheses, and precedence and associativity is specified through \emph{marks} (c.f. Section~\ref{sec:parse-time-disambiguation}). As a delimitation we only consider languages with balanced parentheses. Our approach builds on the following key insights: i) the language of a parse tree is visibly pushdown \cite{alurVisiblyPushdownLanguages2004}, ii) these languages can be encoded as words, and iii) the language of the word encodings of all trees for a language is also visibly pushdown. The last point allows the construction of a visibly pushdown automaton (VPDA) that we examine to solve the static problem for a particular language subclass. Our solution to the dynamic problem, which builds on the first point, is general.

@@

\section{Motivating Examples}
\label{sec:motivation}
In this section, we motivate the need for ambiguous language definitions, where the decision of how to disambiguate a program is taken by the end-user (the programmer) and not the language designer. We motivate our new methodology both for engineering of new domain-specific languages, as well as for the design of existing general-purpose programming languages.

@@

\section{Notation}
\label{sec:prel}

@@

\section{Resolvable Ambiguity} \label{sec:resolvable-definition}

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[sigplan,anonymous,review]{acmart}
%
%% \usepackage{cite}
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

%% \usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
%%                         %% http://ctan.org/pkg/subcaption
%% \usepackage{subfig}
\usepackage{wrapfig}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,automata,calc}
\usepackage{tikz-qtree} % Used for syntax trees
\usepackage{tikz-cd} % Used for commutative diagrams
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

@@

\title{Resolvable Ambiguity}
%% \thanks{This project is financially supported by the Swedish Foundation for Strategic Research (FFL15-0032)} TODO: figure out where to include this
\author{Viktor Palmkvist}
\email{vipa@kth.se}
\orcid{0000-0003-0669-4085}

\author{Elias Castegren}
\email{eliasca@kth.se}
\orcid{0000-0003-4918-6582}

\author{Philipp Haller}
\email{phaller@kth.se}
\orcid{0000-0002-2659-5271}

\author{David Broman}
\email{dbro@kth.se}
\orcid{0000-0001-8457-4105}

\affiliation{%
  \institution{KTH Royal Institute of Technology}
  \streetaddress{100 44}
  \city{Stockholm}
  \state{Sweden}
  }

\renewcommand{\shortauthors}{V. Palmkvist et al.}

%
%
\begin{abstract}
A common standpoint when designing the syntax of programming languages is that
the grammar definition has to be unambiguous.  However, requiring up front unambiguous grammars
can force language designers to make more or less arbitrary choices to disambiguate
the language. In this paper, we depart from the traditional view of
unambiguous grammar design, and enable the detection of ambiguities to be delayed until
parse time, allowing the user of the language to perform the disambiguation. A
natural decision problem follows: given a language definition, can a
user always disambiguate an ambiguous program?  We introduce and
formalize this fundamental problem---called the \emph{resolvable
  ambiguity problem}---and divide it into separate static and dynamic
resolvability problems. We provide solutions to the static problem
for a restricted language class and sketch proofs of soundness and completeness. We also provide a sound
and complete solution to the dynamic problem for a much less
restricted class of languages.
%
The approach is evaluated through two separate case studies,
covering both a large existing programming language, and the
composability of domain-specific languages.
\end{abstract}
%
%
%
\keywords{Syntax, Ambiguity, Grammars, Parsing, Domain-specific languages}

\maketitle              % typeset the header of the contribution

@@

% Motivate why unambiguous parsing is a good (and important) idea.
Ever since the early 60s, it has been known that determining whether a context-free
grammar is ambiguous is undecidable~\cite{cantorAmbiguityProblemBackus1962}. As a consequence,
a large number of restricted grammars have been developed to guarantee
that language definitions are unambiguous. This traditional view of
only allowing unambiguous grammars has been taken for granted as the
only truth: \emph{the} way of how the syntax of a language must be
defined~\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}.
%
However, our recent work~\cite{palmkvistCreatingDomainSpecificLanguages2019} suggests that
carefully designed ambiguous syntax definitions can be
preferable compared to completely unambiguous
definitions.

@@

In this paper we describe a syntax definition formalism where ambiguities can be resolved by grouping parentheses, and precedence and associativity is specified through \emph{marks} (c.f. Section~\ref{sec:parse-time-disambiguation}). As a delimitation we only consider languages with balanced parentheses. Our approach builds on the following key insights: i) the language of a parse tree is visibly pushdown~\cite{alurVisiblyPushdownLanguages2004}, ii) these languages can be encoded as words, and iii) the language of the word encodings of all trees for a language is also visibly pushdown. The last point allows the construction of a visibly pushdown automaton (VPDA) that we examine to solve the static problem for a particular language subclass. Our solution to the dynamic problem, which builds on the first point, is general.

@@

\section{Motivating Examples}\label{sec:motivation}
In this section, we motivate the need for ambiguous language definitions, where the decision of how to disambiguate a program is taken by the end-user (the programmer) and not the language designer. We motivate our new methodology both for engineering of new domain-specific languages, as well as for the design of existing general-purpose programming languages.

@@

\section{Notation}\label{sec:prel}

@@

% TODO: this actually refers to ``Concrete Syntax'' and ``Abstract Syntax Trees'', i.e., not to ``Concrete Syntax Trees''. Not sure how to clarify that
\section{Describing Concrete Syntax and Abstract Syntax Trees}\label{sec:grammars}

This section describes the grammars we use to describe the concrete
syntax and corresponding abstract syntax trees (ASTs) of programs. We
start with context-free grammars and modify them step by step. The
purpose of these modifications is two-fold: we require explicit
descriptions of ASTs, and we require these ASTs to discard some
information from the original parsed program. The discarded
information can then be used for disambiguation, which we describe in
more detail in Section~\ref{sec:tree-language}.

We use the following small language with addition, multiplication,
lists, let-expressions, and numeric literals as a running example:

\begin{center}
  \begin{minipage}[t]{.5\columnwidth}
    \begin{tabular}{@{}l@{\quad}l@{}}
      $E$\quad$->$ & $E$ \texttt{'+'} $E$ \\
      $E$\quad$->$ & $E$ \texttt{'*'} $E$ \\
      $E$\quad$->$ & \texttt{'['} $E_l$ \texttt{']'} \\
      $E$\quad$->$ & \texttt{'let'} \texttt{Ident} \\ % TODO: Ident stands for any identifier, which means that 'Ident' is confusing, but I'm not entirely sure how to present it
      & \texttt{'='} $E$ \texttt{'in'} $E$ \\
      $E$\quad$->$ & \texttt{Int} \\
    \end{tabular}
  \end{minipage}
  \begin{minipage}[t]{.3\columnwidth}
    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
      $E_l$ & $\epsilon$ \\
      $E_l$ & $E$ $E_{l'}$ \\
      \addlinespace
      $E_{l'}$ & $\epsilon$ \\
      $E_{l'}$ & \texttt{','} $E$ $E_{l'}$ \\
    \end{tabular}
  \end{minipage}
\end{center}

\noindent Monospace font, \texttt{'quoted'} or \texttt{otherwise},
represents terminals, $\epsilon$ represents the empty sequence.

First we change the right-hand side of each production to be a regular
expression instead of a sequence (i.e., our grammars are in Extended
Backus-Naur Form, or EBNF) and add (unique) labels to each production.
We use parentheses for grouping, $+$ for union, and $^{*}$ for Kleene
star.

\begin{center}
  \begin{tabular}{@{}l@{\quad$->$\quad}c@{ $:$\quad}l@{}}
    $E$ & $a$ & $E$ \texttt{'+'} $E$ \\
    $E$ & $m$ & $E$ \texttt{'*'} $E$ \\
    $E$ & $l$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
    $E$ & $x$ & \texttt{'let'} \texttt{Ident} \texttt{'='} $E$ \texttt{'in'} $E$ \\
    $E$ & $n$ & \texttt{Int} \\
  \end{tabular}
\end{center}

\noindent Both changes serve to shape our ASTs. For example,
consider the string \texttt{'[1,2]'} with EBNF (left) and without (right):

\begin{center}
  \begin{tabular}{c@{\qquad}c}
    \begin{tikzpicture}
      \tikzset{level 1+/.style={level distance=20pt}}
      \tikzset{frontier/.style={distance from root=40pt}}
      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-4pt) -| (\tikzchildnode)}}}
      \Tree [.{$l$}
        \texttt{[}
          [.{$n$} \texttt{1} ]
          \texttt{,}
          [.{$n$} \texttt{2} ]
          \texttt{]}
      ]
    \end{tikzpicture} &
    \begin{tikzpicture}
      \tikzset{level 1+/.style={level distance=20pt}}
      \tikzset{frontier/.style={distance from root=80pt}}
      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-4pt) -| (\tikzchildnode)}}}
      \Tree [.{$l$}
        \texttt{[}
          [.{$l'$}
            [.{$n$} \texttt{1} ]
            [.{$l''$}
              \texttt{,}
              [.{$n$} \texttt{2} ]
              [.{$l'''$} ] ] ]
          \texttt{]} ]
    \end{tikzpicture}\\
  \end{tabular}
\end{center}

\noindent The extra internal nodes in the right tree would present
problems for disambiguation later.

Next we add grouping parentheses in the form of a pair of terminals
per non-terminal. Syntactically speaking these are equivalent with
adding a production $E \rightarrow \texttt{'('} E \texttt{')'}$,
except they are discarded when constructing the AST. Note the
difference between $()$ and \texttt{'('}\texttt{')'}, the former is
grouping in a regular expression, the latter is a sequence of two
terminals. We also allow discarding terminals that carry no useful
semantic information once the production has been identified. Since
most terminals do \emph{not} carry useful information we denote the
ones to be kept by underlining them:

\begin{center}
  \begin{tabular}{@{}l@{\quad$->$\quad}c@{ $:$\quad}l@{}}
    \multicolumn{3}{l}{Productions} \\
    \midrule
    $E$ & $a$ & $E$ \texttt{'+'} $E$ \\
    $E$ & $m$ & $E$ \texttt{'*'} $E$ \\
    $E$ & $l$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
    $E$ & $x$ & \texttt{'let'} \underline{\texttt{Ident}} \texttt{'='} $E$ \texttt{'in'} $E$ \\
    $E$ & $n$ & \underline{\texttt{Int}} \\
  \end{tabular}
  \begin{tabular}{c} % TODO: figure out how to top align these two
    Grouping \\
    \midrule
    \texttt{'('}$E$\texttt{')'} \\
  \end{tabular}
\end{center}

\noindent We can thus use grouping parentheses and discarded terminals
for disambiguation. The former is often used for operator expressions,
while the latter can be used to, e.g., disambiguate with optional
semicolons. For example, the strings \texttt{'(1+2)*3'} and
\texttt{'[1,2]'} produce these ASTs:

\begin{center}
  \begin{tabular}{c@{\qquad}c}
    \begin{tikzpicture}
      \tikzset{level 1+/.style={level distance=16pt}}
      \tikzset{frontier/.style={distance from root=48pt}}
      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-2pt) -| (\tikzchildnode)}}}
      \Tree [.{$m$}
        [.{$a$}
          [.{$n$} \texttt{1} ]
          [.{$n$} \texttt{2} ] ]
        [.{$n$} \texttt{3} ] ]
    \end{tikzpicture} &
    \begin{tikzpicture}
      \tikzset{level 1+/.style={level distance=16pt}}
      \tikzset{frontier/.style={distance from root=32pt}}
      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-2pt) -| (\tikzchildnode)}}}
      \Tree [.{$l$}
        [.{$n$} \texttt{1} ]
        [.{$n$} \texttt{2} ] ]
    \end{tikzpicture} \\
  \end{tabular}

\end{center}

\noindent Note that most terminals have been discarded from the
ASTs. Note also that we need to keep track of where each node came
from, i.e., which terminal or non-terminal in the grammar it
represents. For example, the $a$ node in the left AST came from the
first $E$ in the right-hand side of the $m$ production. This could,
e.g., be achieved by attaching unique labels to each non-terminal and
kept terminal in a right-hand side, then additionally annotating each
AST node with the corresponding label. However, in most examples this
connection will be obvious, thus we omit these to avoid cluttering the
presentation, and clarify the connection in text if necessary.

Finally we consider precedence and associativity. A common approach is
to create a \emph{precedence ladder} by splitting each non-terminal
into several non-terminals, one per precedence level. Explicitly doing
this is undesireable in our setting, for two reasons:

\begin{enumerate}
\item It requires total precedence, but we wish to allow, e.g.,
  leaving the precedence between operators from different languages
  undefined.
\item Grouping parentheses as described above recurse to the
  \emph{same} non-terminal, which would then not contain productions
  representing lower precedence operators.
\end{enumerate}

\noindent The former suggests that we cannot use precedence ladders
unchanged, and the latter suggests that we need some explicit support
in our grammars; we cannot simply rewrite the grammar without breaking
grouping parentheses.

We thus adopt an approach that coincides with precedence ladders if
precedence is total and transitive, but allows the absence of both
these properties. To simplify later analysis we treat precedence and
associativity as user-facing conveniences to be translated to a
simpler form. We describe the simpler form first, then the
translation, then illustrate both with examples.

We introduce \emph{marks}, annotations on the non-terminals appearing
on the right-hand side of a production. Each mark is a set of
production labels denoting the productions that may \emph{not} appear
in that position. For example, $E_{\{a, x\}}$ may parse as neither the
production labelled $a$, nor the production labelled $x$. It may,
however, parse as a pair of grouping parentheses that contain either of
these productions. We write $E_\emptyset$ as $E$ to reduce clutter.

Translation is then a process of examining the relative precedence of
every pair of productions and the associativity of each production and
adding marks as appropriate. For example, if $E -> m: E \texttt{'*'}
E$ has higher precedence than $E -> a: E \texttt{'+'} E$ we update $m$
to $E -> m: E_{\{a\}} \texttt{'*'} E_{\{a\}}$. If $m$ is additionally
left-associative we update it to $E -> m: E_{\{a\}} \texttt{'*'}
E_{\{a, m\}}$.

% TODO: figure out how to lay these out nicely. The first and third subfigure should fit vertically and be roughly the same height as the second.
\begin{figure*}[t]
  \subcaptionbox{The completed running example.\label{fig:grammar-example}}[.48\textwidth]{
    \begin{tabular}{@{}l@{\quad$->$\quad}c@{ $:$\quad}l@{}}
      \multicolumn{3}{l}{Productions} \\
      \midrule
      $E$ & $a$ & $E_{\{x\}}$ \texttt{'+'} $E_{\{x,a\}}$ \\
      $E$ & $m$ & $E_{\{x,a\}}$ \texttt{'*'} $E_{\{x,a,m\}}$ \\
      $E$ & $l$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
      $E$ & $x$ & \texttt{'let'} \underline{\texttt{Ident}} \texttt{'='} $E$ \texttt{'in'} $E$ \\
      $E$ & $n$ & \underline{\texttt{Int}} \\
    \end{tabular}
    \begin{tabular}{c} % TODO: figure out how to top align these two
      Grouping \\
      \midrule
      \texttt{'('}$E$\texttt{')'} \\
    \end{tabular}
  }
  \subcaptionbox{The running example, translated to EBNF.\label{fig:grammar-example-ebnf}}[.48\textwidth]{
    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
      $E$ & $E_{\{x\}}$ \texttt{'+'} $E_{\{x,a\}}$ \\
      $E$ & $E_{\{x,a\}}$ \texttt{'*'} $E_{\{x,a,m\}}$ \\
      $E$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
      $E$ & \texttt{'let'} \texttt{Ident} \texttt{'='} $E$ \texttt{'in'} $E$ \\
      $E$ & \texttt{Int} \\
      $E$ & \texttt{'('} $E$ \texttt{')'} \\
      \addlinespace
      $E_{\{x\}}$ & $E_{\{x\}}$ \texttt{'+'} $E_{\{x,a\}}$ \\
      $E_{\{x\}}$ & $E_{\{x,a\}}$ \texttt{'*'} $E_{\{x,a,m\}}$ \\
      $E_{\{x\}}$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
      $E_{\{x\}}$ & \texttt{Int} \\
      $E_{\{x\}}$ & \texttt{'('} $E$ \texttt{')'} \\
      \addlinespace
      $E_{\{x,a\}}$ & $E_{\{x,a\}}$ \texttt{'*'} $E_{\{x,a,m\}}$ \\
      $E_{\{x,a\}}$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
      $E_{\{x,a\}}$ & \texttt{Int} \\
      $E_{\{x,a\}}$ & \texttt{'('} $E$ \texttt{')'} \\
      \addlinespace
      $E_{\{x,a,m\}}$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
      $E_{\{x,a,m\}}$ & \texttt{Int} \\
      $E_{\{x,a,m\}}$ & \texttt{'('} $E$ \texttt{')'} \\
    \end{tabular}
  }
  \subcaptionbox{The running example, written in EBNF with a precedence ladder.\label{fig:grammar-example-ebnf-ladder}}[.48\textwidth]{
    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
      $E$ & \texttt{'let'} \texttt{Ident} \texttt{'='} $E$ \texttt{'in'} $E$ \\
      $E$ & $E_1$ \\
      \addlinespace
      $E_1$ & $E_1$ \texttt{'+'} $E_2$ \\
      $E_1$ & $E_2$ \\
      \addlinespace
      $E_2$ & $E_2$ \texttt{'*'} $E_3$ \\
      $E_2$ & $E_3$ \\
      \addlinespace
      $E_3$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
      $E_3$ & \texttt{Int} \\
      $E_3$ & \texttt{'('} $E$ \texttt{')'} \\
    \end{tabular}
  }
  \caption{Completed grammars of the running example.}
\end{figure*}

% TODO: consistency between ``left associative'' and ``left-associative''
Applying the usual precedences to our running example (let $<$
addition $<$ multiplication) as well as associativities
(left-associative multiplication and addition) yields the grammar in
Figure~\ref{fig:grammar-example}\footnote{The observant reader may
  notice that this grammar no longer recognizes, e.g., \texttt{1 + let
    x = 2 in 3}. We discuss this in Section~\ref{sec:low-prec-unary}}.
Note that production $a$ has marks containing $x$ due to precedence
and $a$ due to associativity, and correspondingly $m$ has marks
containing both $x$ and $a$ due to precedence, and $m$ due to
associativity.

To illustrate the concrete syntax in more familiar terms we also show
a translation to normal EBNF grammars in
Figure~\ref{fig:grammar-example-ebnf}. The translation is done by
generating a new non-terminal per uniquely marked non-terminal,
duplicating all productions that are \emph{not} in the mark, and
adding grouping parentheses. This can be compared with
Figure~\ref{fig:grammar-example-ebnf-ladder} which is the same
language, but written more directly as a precedence ladder. Note that
Figure~\ref{fig:grammar-example-ebnf} is equivalent with
Figure~\ref{fig:grammar-example-ebnf-ladder} after inlining all
passthrough productions (e.g. $E -> E_1$).

When describing grammar formalisms a natural question is that of
expressiveness: how expressive are the grammars we present here?
Technically speaking they retain the full expressiveness of
context-free languages; we can trivially translate a CFG by simply
generating unique labels for each production and retaining all
terminals. However, such a translation would benefit from none of our
work as no information is discarded during parsing. We explore the
implications of this in the evaluation, in
Section~\ref{sec:expressiveness}.

We are now ready to describe the language of an AST.

\section{The Language of an Abstract Syntax Tree}\label{sec:tree-language}

Central to our approach is the idea that a single AST can be written
in multiple ways. This means that an AST corresponds to a set of
written programs, i.e., a language. We thus discuss two kinds of
languages: those described by the grammars in
Section~\ref{sec:grammars}, and those consisting of programs that may
parse as a given AST. We refer to the latter as
\emph{AST-languages}. The grammars described in
Section~\ref{sec:grammars} are designed to keep these AST-languages in
the class of visibly pushdown languages
\cite{alurVisiblyPushdownLanguages2004} which have many useful closure
properties. This section informally describes the process of obtaining
an AST-language from an AST and a grammar. We follow the same process
of explanation as Section~\ref{sec:grammars}, this time beginning with
EBNF grammars, then handling each additional modification. Note that
we do not discuss precedence or associativity, we instead handle marks
directly. As a running example we consider the AST obtained by parsing
\texttt{[1, !(2?)]} using the following grammar:

\begin{center}
  \begin{tabular}{@{}l@{\quad$->$\quad}c@{ $:$\quad}l@{}}
    \multicolumn{3}{l}{Productions} \\
    \midrule
    $E$ & $q$ & $E_{\{c\}}$ \texttt{'?'} \\
    $E$ & $c$ & \texttt{'!'} $E$ \\
    $E$ & $l$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) (\texttt{','} $+$ $\epsilon$) \texttt{']'} \\
    $E$ & $n$ & \underline{\texttt{Int}} \\
  \end{tabular}
  \begin{tabular}{c} % TODO: figure out how to top align these two
    Grouping \\
    \midrule
    \texttt{'('}$E$\texttt{')'} \\
  \end{tabular}
\end{center}

When parsing using a normal EBNF grammar, without discarding any
information, a given AST can only be parsed in exactly one way, namely
the flattening of the tree. In this case, the AST (which is then more
like a parse tree) looks as on the left (using $g$ for grouping
parentheses), and the AST-language is recognized by the pushdown
automaton on the right (though we have no need of the stack yet):

\begin{center}
  \begin{tabular}{cc}
    \begin{tikzpicture}
      \tikzset{level 1+/.style={level distance=16pt}}
      \tikzset{frontier/.style={distance from root=80pt}}
      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-2pt) -| (\tikzchildnode)}}}
      \Tree [.{$l$}
        \texttt{[}
        [.{$n$} \texttt{1} ]
        \texttt{,}
        [.{$c$}
          \texttt{!}
          [.{$g$}
            \texttt{(}
            [.{$q$} [.{$n$} \texttt{2} ] \texttt{?} ]
            \texttt{)} ] ]
        \texttt{]} ]
    \end{tikzpicture} &
    \begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=1cm,transform shape]
      \tikzset{every state/.style={minimum size=12pt}}
      \node[state,initial] (1) {};
      \node[state] (2) [right of=1] {};
      \node[state] (3) [below of=2] {};
      \node[state] (4) [left of=3] {};
      \node[state] (5) [left of=4] {};
      \node[state] (6) [below of=5] {};
      \node[state] (7) [right of=6] {};
      \node[state] (8) [right of=7] {};
      \node[state] (9) [below of=8] {};
      \node[state,accepting] (10) [left of=9] {};

      \path (1) edge node{\texttt{[}} (2);
      \path (2) edge node{\texttt{1}} (3);
      \path (3) edge[above] node{\texttt{,}} (4);
      \path (4) edge[above] node{\texttt{!}} (5);
      \path (5) edge[left] node{\texttt{(}} (6);
      \path (6) edge node{\texttt{2}} (7);
      \path (7) edge node{\texttt{?}} (8);
      \path (8) edge node{\texttt{)}} (9);
      \path (9) edge[above] node{\texttt{]}} (10);
    \end{tikzpicture}
  \end{tabular}
\end{center}

\noindent When ASTs additionally discard some terminals the
AST-language is no longer merely a flattening of the tree; instead it
instead has to represent all possible discarded sequences of
terminals. In some cases this yields no difference, e.g., production
$q$ discards exactly one terminal \texttt{'?'}. In other cases there
is a small difference, e.g., the end of production $l$ may discard
either \texttt{',]'} or merely \texttt{']'}. Our running example thus
yields:

\begin{center}
  \begin{tabular}{c@{\qquad}c}
    \begin{tikzpicture}
      \tikzset{level 1+/.style={level distance=16pt}}
      \tikzset{frontier/.style={distance from root=80pt}}
      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-2pt) -| (\tikzchildnode)}}}
      \Tree [.{$l$}
        [.{$n$} \texttt{1} ]
        [.{$c$}
          [.{$g$}
            [.{$q$} [.{$n$} \texttt{2} ] ] ] ] ]
    \end{tikzpicture} &
    \begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=1cm,transform shape]
      \tikzset{every state/.style={minimum size=12pt}}
      \node[state,initial] (1) {};
      \node[state] (2) [right of=1] {};
      \node[state] (3) [right of=2] {};
      \node[state] (4) [right of=3] {};
      \node[state] (5) [right of=4] {};
      \node[state] (6) [below of=5] {};
      \node[state] (7) [left of=6] {};
      \node[state] (8) [left of=7] {};
      \node[state] (9) [left of=8] {};
      % TODO: perform the first line break one step later, then have only the accepting state on the lowest line.
      \node[state] (11) [left of=9] {};
      \node[state,accepting] (10) [below of=11] {};

      \path (1) edge node{\texttt{[}} (2);
      \path (2) edge node{\texttt{1}} (3);
      \path (3) edge node{\texttt{,}} (4);
      \path (4) edge node{\texttt{!}} (5);
      \path (5) edge node{\texttt{(}} (6);
      \path (6) edge[above] node{\texttt{2}} (7);
      \path (7) edge[above] node{\texttt{?}} (8);
      \path (8) edge[above] node{\texttt{)}} (9);
      \path (9) edge node{\texttt{]}} (10);
      \path (9) edge[above] node{\texttt{,}} (11);
      \path (11) edge[left] node{\texttt{]}} (10);
    \end{tikzpicture}
  \end{tabular}
\end{center}

% TODO: I question the clarity of this paragraph. The idea is simple, but I'm not sure I'm conveying it properly
\noindent In general, the discarded sequence of terminals between two
retained points in a production must be described by a regular
language since all non-terminals are retained. For example, the AST
obtained by parsing \texttt{'[]'} consists of a single node $l$ with
no children. Its AST-language is described by the regular expression
$\texttt{'['} (\texttt{','} + \epsilon) \texttt{']'}$ since the other
sub-expressions in the production contain non-terminals (which must be
retained) that are not present in the AST.

Next we take discarded grouping parentheses into account. The AST now
discards the $g$ node, and the automaton must allow inserting an
arbitrary (balanced) amount of parentheses around each node. We ensure
this by pushing to the stack when recognizing a \texttt{'('} and
popping on \texttt{')'}. To ensure that parentheses pair correctly
around each node we use a distinct stack symbol for each node. These
stack symbols are represented by letters below. To reduce clutter we
write $\circ \xrightarrow{+a} \circ$ as shorthand for an edge that
recognizes \texttt{'('} and pushes $a$ on the stack, while $\circ
\xrightarrow{-a} \circ$ correspondingly recognizes \texttt{')'} and
pops $a$ from the stack\footnote{This works for our example since we
  have only a single form of grouping parentheses, otherwise we would
  have to be explicit.}.

\begin{center}
  \begin{tabular}{c@{\qquad}c}
    \begin{tikzpicture}
      \tikzset{level 1+/.style={level distance=16pt}}
      \tikzset{frontier/.style={distance from root=64pt}}
      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-2pt) -| (\tikzchildnode)}}}
      \Tree [.{$l$}
        [.{$n$} \texttt{1} ]
        [.{$c$}
          [.{$q$} [.{$n$} \texttt{2} ] ] ] ]
    \end{tikzpicture} &
    \begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=1cm,transform shape]
      \tikzset{every state/.style={minimum size=12pt}}
      \node[state,initial] (1) {};
      \node[state] (2) [right of=1] {};
      \node[state] (3) [right of=2] {};
      \node[state] (4) [right of=3] {};
      \node[state] (5) [right of=4] {};
      \node[state] (7) [below of=4] {};
      \node[state] (8) [left of=7] {};
      % TODO: perform the first line break one step later, then have only the accepting state on the lowest line.
      \node[state] (11) [below of=1] {};
      \node[state,accepting] (10) [below of=11] {};

      \path (1) edge node{\texttt{[}} (2);
      \path (2) edge node{\texttt{1}} (3);
      \path (3) edge node{\texttt{,}} (4);
      \path (4) edge node{\texttt{!}} (5);
      \path (5) edge node{\texttt{2}} (7);
      \path (7) edge[above] node{\texttt{?}} (8);
      \path (8) edge node{\texttt{]}} (10);
      \path (8) edge[above] node{\texttt{,}} (11);
      \path (11) edge[left] node{\texttt{]}} (10);

      \path (1) edge[loop above] node{$+a$} (1);
      \path (10) edge[loop left] node{$-a$} (1);

      \path (2) edge[loop above] node{$+b$} (2);
      \path (3) edge[loop above] node{$-b$} (3);
